Direct Entropy Measurement in a Mesoscopic Quantum System

### 2018-03-02 paper/draft.tex
cleaning up
mostly text edits. few small figure changes.

@@

@@

\title{Direct Entropy Measurement in a Mesoscopic Quantum System}
\author{Nikolaus Hartman}
\email{nik.hartman@gmail.com}
\thanks{The raw data and Python code used for all figures and analysis can be found at \url{https://github.com/nikhartman/spin_entropy}}
	\affiliation{Stewart Blusson Quantum Matter Institute, University of British Columbia, Vancouver, British Columbia, V6T1Z4, Canada}
	\affiliation{Department of Physics and Astronomy, University of British Columbia, Vancouver, British Columbia, V6T1Z1, Canada}
\author{Christian Olsen}
	\affiliation{Stewart Blusson Quantum Matter Institute, University of British Columbia, Vancouver, British Columbia, V6T1Z4, Canada}
	\affiliation{Department of Physics and Astronomy, University of British Columbia, Vancouver, British Columbia, V6T1Z1, Canada}
\author{Silvia Folk}
	\affiliation{Stewart Blusson Quantum Matter Institute, University of British Columbia, Vancouver, British Columbia, V6T1Z4, Canada}
	\affiliation{Department of Physics and Astronomy, University of British Columbia, Vancouver, British Columbia, V6T1Z1, Canada}
\author{Mohammad Samani}
	\altaffiliation{The Hospital for Sick Children, Toronto, ON, Canada}
	\altaffiliation{Fields Institute for Research in Mathematical Sciences, Toronto, ON, Canada}
	\affiliation{Stewart Blusson Quantum Matter Institute, University of British Columbia, Vancouver, British Columbia, V6T1Z4, Canada}
	\affiliation{Department of Physics and Astronomy, University of British Columbia, Vancouver, British Columbia, V6T1Z1, Canada}
\author{Saeed Fallahi}
	\affiliation{Department of Physics and Astronomy, Purdue University}
	\affiliation{Station Q Purdue, Purdue University}
	\affiliation{Birck Nanotechnology Center, Purdue University, West Lafayette, Indiana, USA}
\author{Geoffrey C. Gardner}
	\affiliation{Station Q Purdue, Purdue University}
	\affiliation{Birck Nanotechnology Center, Purdue University, West Lafayette, Indiana, USA}
    	\affiliation{School of Materials Engineering, Purdue University, West Lafayette, Indiana, USA}
\author{Michael Manfra}
	\affiliation{Department of Physics and Astronomy, Purdue University}
	\affiliation{Station Q Purdue, Purdue University}
	\affiliation{Birck Nanotechnology Center, Purdue University, West Lafayette, Indiana, USA}
	\affiliation{School of Electrical and Computer Engineering,  Purdue University, West Lafayette, Indiana, USA}
    	\affiliation{School of Materials Engineering, Purdue University, West Lafayette, Indiana, USA}
\author{Joshua Folk}
\email{jfolk@physics.ubc.ca}
	\affiliation{Stewart Blusson Quantum Matter Institute, University of British Columbia, Vancouver, British Columbia, V6T1Z4, Canada}
	\affiliation{Department of Physics and Astronomy, University of British Columbia, Vancouver, British Columbia, V6T1Z1, Canada}
\date{\today}

@@

%%%%%%% intro material %%%%%%%%%
\textbf{The thermodynamic properties of electronic systems offer important insights into the nature of their ground states, and can probe exotic quasiparticles that may emerge due to interactions or non-trivial topology.  Systems that are difficult to clearly identify through standard conductance measurements may be characterized more precisely with a thermodynamic measurement. For example, the purported non-Abelian exchange statistics of Moore-Read quasiparticles in the $\nu = \frac{5}{2}$ fractional quantum Hall state, or of Majorana quasiparticles in a topological superconductor, could be clearly distinguished from the Abelian case by an entropy measurement \cite{Cooper2009, Smirnov2015}.  Such measurements are typically based on bulk properties, like heat capacity, that are clearly observed in macroscopic samples but are immeasurably small in systems that may consist of only a few electrons.  Here, we develop a mesoscopic circuit to make direct entropy measurements of the first, second, and third electron ground states in a GaAs quantum dot.  This system is chosen for having well-known entropy based on simple spin states\cite{Tarucha1996, Ciorga2000, Duncan2000, Lindemann2002, Potok2003, Hofmann2016}.  The precision of this technique, quantifying the entropy of a single spin-$\frac{1}{2}$ to within 5\% of the expected value ($k_B \ln{2}$), shows its future potential for probing more exotic systems, from Majorana states to non-Fermi liquid Kondo states, that are clearly identified by their residual low-temperature entropy\cite{Alkurtass2016}.}

%%% figure 1 %%%
\begin{figure}[!]
        \includegraphics[width=1.0\columnwidth]{../figures/figure_1_annotated.pdf}
        \caption{\label{fig:fig1}(a) Scanning electron micrograph of a device similar to the one measured, showing electrostatic gates (gold) used to define a circuit in a 2D electron gas (2DEG). Grey gates are grounded. Squares show the locations of ohmic contacts to the 2DEG. An electron reservoir is formed between the parallel gates running vertically down the image, with a quantum dot coupled to the right side. The reservoir can be heated by driving current through the quantum point contact on the left.  Electrons tunneling between the reservoir and dot are measured with a capacitively-coupled charge sensor. (b) and (c) Simulated charge sensor signal for a transition from $N-1 \rightarrow N$ electrons at two temperatures ($T_{Red} > T_{Blue}$), showing two possible cases for $\frac{\partial S}{\partial N}$.}
\end{figure}

@@

The Maxwell relation in Eq.~\ref{eqn:max} forms the basis of two theoretical proposals to measure non-Abelian exchange of quasiparticles in the the $\nu = \frac{5}{2}$ state, by detecting a change in quasiparticle chemical potential with temperature, which is proportional to the entropy derivative, $\frac{\partial S}{\partial N}$ \cite{Cooper2009,Ben-Shach2013}. Applying the language of quantum dots to Eq.~\ref{eqn:max}, the entropy difference between the $N-1$ to $N$ quantum dot ground states ($\Delta S_{N-1\rightarrow N}$ for $\Delta N=1$) is measured through the shift with temperature in the electrochemical potential, $\mu_N$, needed to add the $N$th electron to the dot; the gate voltage at which the $N-1$ to $N$ transition is observed is  proportional to $\mu_{N}$.

The mesoscopic circuit shown in Fig.~\ref{fig:fig1}a uses electrostatic gates to realize an electron reservoir (central region; hereafter ``reservoir") in thermal and diffusive equilibrium with a few-electron quantum dot (``dot") coupled to its right side.  The reservoir temperature, $T$, could be rapidly (up to at least \SI{100}{\hertz}) increased above the mixing chamber temperature, $T_{MC}$, by Joule heating through the QPC (``heater") on the left side.  The occupation of the dot was measured using an adjacent quantum point contact (``charge sensor") \cite{Field1993, Staring2007, Thierschmann2015} [change these references/add additional].  Applying a more positive voltage to the gate adds the $N$th electron to the dot when $\mu_{N}$ drops below the Fermi level of the reservoir, $E_F$. Figure~\ref{fig:fig1}b illustrates such a transition \textemdash seen as a step in the charge sensor conductance \textemdash that is thermally broadened by the reservoir temperature.  The gate voltage corresponding to the midpoint of the transition, $V_{mid}$, marks the electrochemical potential at which the probabilities of finding $N-1$ or $N$ electrons on the dot are equal.  When $\mu_N$ shifts with temperature, $V_{mid}$ also shifts (Fig.~\ref{fig:fig1}c); it is the measurement of this signature that forms the basis of our experiment.

The relation between entropy and any temperature-induced shift in $\mu_{N}$ with respect to $E_F$ can be understood through detailed balance.  The $N$th electron tunnels back and forth between an $N-1$-electron dot and reservoir when there are available states in both the dot and reservoir, that is, when $\mu_{N}$ is within thermal broadening ($\sim k_B T$) of $E_F$.  The tunnel rates into and out of the dot, $\Gamma_{in}=\Gamma_{N-1\rightarrow N}$ and $\Gamma_{out}=\Gamma_{N\rightarrow N-1}$, depend on degeneracy of the $N-1$ and $N$ ground states. Consider, for example, the first electron tunneling into an empty dot: both spin-up and -down states are available on the dot.  Tunneling out, the electron (whether spin-up or -down) must tunnel into reservoir states with the same spin, that is, only half of the empty reservoir states are available.  As a result, $\Gamma_{in} = 2\Gamma_{out}$ when $\mu_{N}=E_F$.  Extending this logic to a general case gives the result derived from detailed balance \cite{Beenakker1991, Gustavsson2009}, 
%
\begin{align}
	\Gamma_{in} &=  g_{N} \Gamma f(E_F - \mu_{N}) \nonumber \\
	\Gamma_{out} &= g_{N-1} \Gamma [1 - f(E_F - \mu_{N})] \label{eqn:rates}
\end{align}
%
where $\Gamma$ is a coupling constant, $f$ is the Fermi function, and $g_{N(N-1)}$ is the degeneracy of the $N(N-1)$ electron state.

The equal probability condition for $N-1$- and $N$-electrons at $V_{mid}$ corresponds to the incoming and outgoing rates being equal. From Eq.~\ref{eqn:rates}$, \Gamma_{in} = \Gamma_{out}$ corresponds to $\mu_{N} = E_F$ only in the case $g_{N-1}=g_{N}$, but to $\mu_{N} > E_F$ when $g_{N-1} > g_{N}$ and to $\mu_{N} < E_F$ when $g_{N-1} < g_{N}$. The ratio $g_{N}/g_{N-1}$ is given by the shift in $V_{mid}$ ($\propto \mu_{N}$) with temperature, giving $\Delta S_{N-1\rightarrow N}$ via the Boltzmann entropy, $S_{N}=k_{B} \ln{g_N}$. Previous works have explored the relationship between tunnel rates and degeneracy using time-resolved transport spectroscopy and coupling of quantum dots to atomic force microscope oscillations \cite{Cockins2010, Bennett2010, Beckel2014, Hofmann2016}. The approach presented here illustrates the thermodynamic equivalent of the rate measurement, and extends entropy measurements to a wide set of applications where tunneling processes may not be possible to observe in real-time.

%%%%%%%%%%

%%% figure 2 %%%
\begin{figure}[!]
        \includegraphics[width=1.0\columnwidth]{../figures/figure_2.pdf}
        \caption{\label{fig:fig2}(a) Charge sensor data for $N=0 \rightarrow 1$ at two temperatures set by DC current through the QPC heater. (b) Transition width, $\Theta$, was linear in $T_{MC}$ above 100 mK, for $I_{heat}=0$. Lever arm $\alpha$ is calculated by fitting a straight line to this region.  (c) Lock-in measurement of $\delta g_{sens}$ with $\delta T = \SI{32}{\milli\kelvin}$. Fits to $\delta g_{sens}$ are shown with $\epsilon$ as a free parameter (solid) or fixed at $\epsilon=0$ (dashed)  (d) $\Theta$ grows with DC current through the QPC heater. A fit to $T = a\sqrt{T^{2}_{MC}+b I_{heat}R_{QPC}}$ is used to convert between $I_{heat}$ and $\delta T$, with $R_{QPC} = \SI{20}{\kilo\ohm}$ \cite{Mittal1996} (e)  Entropy measurements were independent of the magnitude of $I_{heat}$ oscillations. Top axis illustrates the corresponding magnitude of $\delta T$, while right axes shows the entropy signal converted to a gate voltage shift.}
\end{figure}

To proceed with the entropy measurement, the quantum dot was tuned such that the source was weakly tunnel coupled to the reservoir with the drain closed.  The conductance of the charge sensor was tuned to $g_{sens}{\sim}e^2/h$, where it was most sensitive to charge on the dot.  The addition of the first electron to the dot was marked by a decrease in $g_{sens}$ that is well fit by the expected form for a thermally-broadened transition:
%
\begin{align}
\label{eqn:g}
        g_{sens}(V,\Theta) &= G_0 \tanh\left(\frac{V-V_{mid}(\Theta)}{2\Theta}\right)  \\
                        &\quad + G_1\left[V-V_{mid}(\Theta)\right] + G_2 \nonumber
\end{align}
%
where $G_0$ represents the sensor sensitivity, $\Theta = \frac{k_B T}{\alpha e}$ is the thermal broadening (expressed in units of gate voltage), $\alpha\equiv d \mu_{N}/d V_G$ is the lever arm, and $G_1$ reflects the cross capacitance between the charge sensor and plunger gate. Fig.~\ref{fig:fig2}a shows two such transition curves with the high temperature curve being thermally broadened by driving a current, $I_{heat}$ through the QPC heater. With no Joule heating through the heater, $\Theta$ followed $T_{MC}$ down to \SI{<100}{\milli\kelvin} (Fig.~\ref{fig:fig2}b), validating the approximation of thermal broadening used throughout this experiment.

The data in Fig.~\ref{fig:fig2}c, and corresponding fits, offer an example of how $\Delta S_{0\rightarrow 1}$ was measured across the $0 \rightarrow 1$ electron transition.  A comparison of the curves in Fig.~\ref{fig:fig2}a shows that $g_{sens}$ drops as temperature rises immediately before the midpoint of the transition ($V_G-V_{mid}<0$), whereas $g_{sens}$ rises with temperature for $V_G-V_{mid}>0$.   Oscillating the reservoir temperature by $\delta T$ using an AC current through the heater while measuring oscillations in the charge sensor signal, $\delta g_{sens}$, with a lock-in amplifier (see Methods) yields the characteristic peak-dip structure seen in Fig.~\ref{fig:fig2}c. The expected lineshape of such a curve is given by $\delta g = \frac{\partial g}{\partial T} \delta T$, where $g(V;T)$ is defined by Eq.~\ref{eqn:g} and $\delta T$ is determined by the RMS value of $I_{heat}$:
%
\begin{align}
\label{eqn:dg}
        \delta g_{sens}(V_G, \Theta) &\propto -\delta T \left[ \frac{V_G-V_{mid}(\Theta)}{2\Theta} + \frac{\epsilon}{2} \right]\times \\
        				      &\quad\cosh^{-2}\left(\frac{V_G-V_{mid}(\Theta)}{2\Theta}\right) + const. \nonumber
\end{align}
%
where $\epsilon=\frac{\partial V_{mid}}{\partial \Theta}$. Using the lever arm to convert $V_{mid}$ to electrochemical potential, $\epsilon$ is simply the change in entropy (scaled by $k_B$) through Eq.~\ref{eqn:max}:
%
% 
%
\begin{align}
\label{eqn:epsilon}
        \epsilon &= \frac{\partial V_{mid}}{\partial \Theta} = 
        \frac{1}{k_B} \frac{\partial \mu}{\partial T} = 
        -\frac{1}{k_B} \Delta S_{N-1\rightarrow N}
\end{align}
%

From Eqs.~\ref{eqn:dg} and \ref{eqn:epsilon}, $\delta g_{sens}(V_G)$ would be perfectly antisymmetric around $V_{mid}$ if $\Delta S_{N-1\rightarrow N}$ were zero, whereas the data in Fig.~\ref{fig:fig2}c are clearly not antisymmetric. A best fit to Eqs.~\ref{eqn:dg} and \ref{eqn:epsilon} yields $\Delta S_{0\rightarrow 1}=(1.02 \pm 0.01) k_B \ln{2}$. For the $0 \rightarrow 1$ electron transition at zero magnetic field, one expects the following: $S_0$ for the empty state should be $k_B \ln{1}=0$ whereas for the spin degenerate 1-electron state one expects $S_1=k_B\ln{2}$, giving $\Delta S_{0\rightarrow 1}\equiv S_1 - S_0 =k_B\ln{2}$, matching the very well with the measured value.

Figure~\ref{fig:fig2}e shows that $\Delta S$, determined from fits analogous to those in Fig.~\ref{fig:fig2}c, remains constant over a range of $\delta T$ ($I_{heat}$ AC). The calibration for $\delta T$ with respect to $I_{heat}$ is shown in Figure~\ref{fig:fig2}d; a DC heater current was applied and fits to \label{eqn:g} yield $\delta T$ through $\Theta$; the calibration holds for AC currents when referring to the RMS value of $I_{heat}$ AC, as we do throughout this discussion. An important point is illustrated in Figure~\ref{fig:fig2}e; the change in entropy, $\Delta S$, is determined entirely by the fit parameter, $\epsilon$, with no further calibration necessary.

%%% figure 3 %%%
\begin{figure}
        \includegraphics[width=1.0\columnwidth]{../figures/figure_3.pdf}
        \caption{\label{fig:fig3}(a) Changes in entropy for $N=0 \rightarrow 1$ and $2 \rightarrow 3$ transitions, overlaid to highlight similar behaviors.  Each data point corresponds to a single $\delta g_{sens}(V_G)$ fit; multiple scans are carried out at various parallel magnetic fields.   (b) and (c) Characteristic $\delta g_{sens}$ traces from which the data in (a) were extracted. The two data points corresponding to (b) and (c) are shown as large markers in (a). (e) Bias spectroscopy data for the $N=0 \rightarrow 1$ transition. Dashed line at $V_{SD}$ = \SI{700}{\micro\electronvolt} shows where data in (f) are taken. (f) Fixed bias data showing fits to Zeeman splitting of the ground state (dashed lines) from which $g = -0.42 \pm 0.01$ is extracted.}
\end{figure}

>>>>

% manfra comments:
%         didn't understand how the measurement was done. thought we were measuring Vmid and seeing it shift.  
%         wanted to know more about how this technique would actually be implemented to realize one of the more exciting proposals.

% mohammad comments:
% 	    Figure 2 -- (c) make more clear how delta T was obtained
%         make more clear how the 5/2 measurement relates to what was done here
%         gate-tuned -- change this phrase in the methods
%         I_heat looks like it shorts the sample in Figure 1a
%         Nature Methods -- suggestion for journal

@@

\newcommand{\ket}[1]{\ensuremath{\left|#1\right\rangle}}

@@

\title{Direct Entropy Measurement in a Mesoscopic Quantum System}
\author{Nikolaus Hartman}
\email{nik.hartman@gmail.com}
\thanks{The raw data and Python code used for all figures and analysis can be found at \url{https://github.com/nikhartman/spin_entropy}}
	\affiliation{Stewart Blusson Quantum Matter Institute, University of British Columbia, Vancouver, British Columbia, V6T1Z4, Canada}
	\affiliation{Department of Physics and Astronomy, University of British Columbia, Vancouver, British Columbia, V6T1Z1, Canada}
\author{Christian Olsen}
	\affiliation{Stewart Blusson Quantum Matter Institute, University of British Columbia, Vancouver, British Columbia, V6T1Z4, Canada}
	\affiliation{Department of Physics and Astronomy, University of British Columbia, Vancouver, British Columbia, V6T1Z1, Canada}
\author{Silvia Folk}
	\affiliation{Stewart Blusson Quantum Matter Institute, University of British Columbia, Vancouver, British Columbia, V6T1Z4, Canada}
	\affiliation{Department of Physics and Astronomy, University of British Columbia, Vancouver, British Columbia, V6T1Z1, Canada}
\author{Mohammad Samani}
	\altaffiliation{The Hospital for Sick Children, Toronto, ON, Canada}
	\altaffiliation{Fields Institute for Research in Mathematical Sciences, Toronto, ON, Canada}
	\affiliation{Stewart Blusson Quantum Matter Institute, University of British Columbia, Vancouver, British Columbia, V6T1Z4, Canada}
	\affiliation{Department of Physics and Astronomy, University of British Columbia, Vancouver, British Columbia, V6T1Z1, Canada}
\author{Saeed Fallahi}
	\affiliation{Department of Physics and Astronomy, Purdue University, West Lafayette, Indiana, USA}
	\affiliation{Station Q Purdue, Purdue University, West Lafayette, Indiana, USA}
	\affiliation{Birck Nanotechnology Center, Purdue University, West Lafayette, Indiana, USA}
\author{Geoffrey C. Gardner}
	\affiliation{Station Q Purdue, Purdue University, West Lafayette, Indiana, USA}
	\affiliation{Birck Nanotechnology Center, Purdue University, West Lafayette, Indiana, USA}
    	\affiliation{School of Materials Engineering, Purdue University, West Lafayette, Indiana, USA}
\author{Michael Manfra}
	\affiliation{Department of Physics and Astronomy, Purdue University, West Lafayette, Indiana, USA}
	\affiliation{Station Q Purdue, Purdue University, West Lafayette, Indiana, USA}
	\affiliation{Birck Nanotechnology Center, Purdue University, West Lafayette, Indiana, USA}
	\affiliation{School of Electrical and Computer Engineering,  Purdue University, West Lafayette, Indiana, USA}
    	\affiliation{School of Materials Engineering, Purdue University, West Lafayette, Indiana, USA}
\author{Joshua Folk}
\email{jfolk@physics.ubc.ca}
	\affiliation{Stewart Blusson Quantum Matter Institute, University of British Columbia, Vancouver, British Columbia, V6T1Z4, Canada}
	\affiliation{Department of Physics and Astronomy, University of British Columbia, Vancouver, British Columbia, V6T1Z1, Canada}
\date{\today}

@@

%%%%%%% intro material %%%%%%%%%
\textbf{The thermodynamic properties of electronic systems offer important insights into the nature of their ground states, and can probe exotic quasiparticles that may emerge due to interactions or non-trivial topology.  Systems that are difficult to clearly identify through standard conductance measurements may be characterized more precisely with a thermodynamic measurement. For example, the purported non-Abelian exchange statistics of Moore-Read quasiparticles in the $\nu = \frac{5}{2}$ fractional quantum Hall state, or of Majorana quasiparticles in a topological superconductor, could be clearly distinguished from the Abelian case by an entropy measurement \cite{Cooper2009, Ben-Shach2013, Smirnov2015}.  
Such measurements are typically based on bulk properties, like heat capacity, that are easily observed in macroscopic samples but are immeasurably small in systems that may consist of only a few electrons.  Here, we develop a mesoscopic circuit to make direct entropy measurements of single- or few-quasiparticle states---in this case, the first, second, and third electron ground states in a GaAs quantum dot.  This system is chosen for having well-known entropy based on simple spin states\cite{Tarucha1996, Ciorga2000, Duncan2000, Lindemann2002, Potok2003, Hofmann2016}.  The precision of this technique, quantifying the entropy of a single spin-$\frac{1}{2}$ to within 5\% of the expected value ($k_B \ln{2}$), shows its future potential for probing more exotic systems, from Majorana states to non-Fermi liquid Kondo states, that are clearly identified by their residual low-temperature entropy\cite{Alkurtass2016}.}

%%% figure 1 %%%
\begin{figure}[!]
        \includegraphics[width=1.0\columnwidth]{../figures/figure_1_annotated.pdf}
        \caption{\label{fig:fig1} \textbf{Entropy measurement protocol.}  (a) Scanning electron micrograph of a device similar to the one measured. Electrostatic gates (gold) define the circuit in a 2D electron gas (2DEG), with grey gates grounded. Squares indicate ohmic contacts to the 2DEG.  The temperature of the electron reservoir in the middle (red) is oscillated using AC current, $I_{heat}$, at frequency $f_{heat}$ through the quantum point contact (QPC) on the left.  A portion of the $5~\mu$m-wide reservoir has been removed here for clarity.  The occupation of the quantum dot, tunnel coupled to the right side the reservoir, is tuned by $V_p$ and monitored by $I_{sens}$ through the charge sensor QPC. $I_{sens}$ is split into DC and AC components, the latter being measured by a lock-in amplifier at $2f_{heat}$.  (b) and (c) Simulated DC charge sensor signal, $G_{sens}$, for a transition from $N-1 \rightarrow N$ electrons at two temperatures ($T_{Red} > T_{Blue}$), showing two possible cases for $\frac{\partial S}{\partial N}$. Insets show the corresponding difference, $\delta G_{sens}$, between hot and cold curves.}
\end{figure}

@@

The Maxwell relation in Eq.~\ref{eqn:max} forms the basis of two theoretical proposals to measure non-Abelian exchange of quasiparticles in  the $\nu = \frac{5}{2}$ state via their entropy
%, quantified by the change in quasiparticle chemical potential with temperature
\cite{Cooper2009,Ben-Shach2013}.  Ref.~\onlinecite{Ben-Shach2013} proposes a strategy by which quasiparticle entropy could be deduced from the temperature-dependent shift of charging events on a local disorder potential, a thermodynamic equivalent of the measurements that established the $e/4$ quasiparticle charge\cite{Venkatachalam2011}.   As a proof of principle in the viability of measuring localized quasiparticle entropy, and a demonstration of the high accuracy achievable by this technique, we investigate a well-understood system with localized fermions: a few-electron GaAs quantum dot. The entropies of the first three electron states in the dot are measured by the temperature-dependent charging scheme laid out in Ref.~\onlinecite{Ben-Shach2013}.   Applying the language of quantum dots to Eq.~\ref{eqn:max}, the entropy difference between the $N-1$ and $N$ electron ground states ($\Delta S_{N-1\rightarrow N}$ for $\Delta N=1$) is measured via the shift with temperature in the electrochemical potential, $\mu_N$, needed to add the $N$th electron to the dot.

The measurement relies on the mesoscopic circuit shown in Fig.~\ref{fig:fig1}a, using electrostatic gates to realize an electron reservoir (hereafter ``reservoir") in thermal and diffusive equilibrium with a few-electron quantum dot (``dot") coupled to its right side.  The occupation of the dot is tuned with the plunger gate voltage, $V_p$, and measured using an adjacent quantum point contact (``charge sensor") \cite{Field1993, Staring2007, Thierschmann2015}.  Applying more positive $V_p$ lowers $\mu_N$,  bringing the $N$th electron into the dot when $\mu_{N}$ drops below the Fermi level of the reservoir, $E_F$. The reservoir temperature, $T$, can be increased above the mixing chamber temperature, $T_{MC}$, by Joule heating from current $I_{heat}$ through the quantum point contact (``heater") on the left side.   Charge transitions on the dot appear as steps in the charge sensor conductance, $G_{sens}(V_p)$, thermally broadened by the reservoir temperature (Figs.~\ref{fig:fig1}b and c).  The gate voltage corresponding to the midpoint of the transition, $V_{mid}$, marks the electrochemical potential at which the probabilities of finding $N-1$ and $N$ electrons on the dot are equal.

When $\mu_N$ shifts with temperature (Eq.~\ref{eqn:max}), $V_{mid}$ also shifts; it is the shift in $V_{mid}$ with temperature that forms the basis of our experiment (Fig.~\ref{fig:fig1}c).  In practice, charge noise limits the accuracy to which $V_{mid}$ can be measured, so the measurement is done with a lock-in amplifier, oscillating the temperature using an AC $I_{heat}$ and measuring resultant oscillations in $G_{sens}$.  As seen in the insets of Figs.~\ref{fig:fig1}b and c, the lineshape of $\delta G_{sens}$ is perfectly antisymmetric when $\partial S/\partial N=0$, but asymmetric when $\partial S/\partial N \neq 0$.

The temperature-induced shift in the dot chemical potential with respect to reservoir $E_F$ can also be understood in terms of detailed balance.  At $V_{mid}$, with equal probabilities for $N$ and $N-1$ electrons on the dot, the tunnel rates $\Gamma_{in}=\Gamma_{N-1\rightarrow N}$ and $\Gamma_{out}=\Gamma_{N\rightarrow N-1}$ must also be equal. These rates depend on the number of available states in the tunneling process, and therefore on the degeneracies of the $N-1$ and $N$ ground states \cite{Beenakker1991, Gustavsson2009}.  The condition $\Gamma_{in} = \Gamma_{out}$ leads to a simple relationship between the degeneracies, $d_{N-1}$ and $d_{N}, of the $N-1$ and $N$ electron states and the thermally broadened Fermi function $f$: $d_{N-1}/d_{N}=f/(1-f)$. Using the Boltzman entropy, $S_{N}=k_{B} \ln{d_N}$, and the form of the Fermi function, this relationship becomes $\Delta S_{N-1\rightarrow N}= (\mu_{N}-E_F)/T$, clearly demonstrating the connection between entropy, temperature, and the shift in $\mu_N$ relative to $E_F$ at $V_{mid}$. Previous experiments have explored the relationship between tunnel rates and degeneracy using time-resolved transport spectroscopy and by coupling quantum dots to atomic force cantilever oscillations \cite{Cockins2010, Bennett2010, Beckel2014, Hofmann2016}. The approach presented here is a thermodynamic equivalent of these, and extends entropy measurements to a much wider set of applications where tunnelling processes may not be possible to observe in real-time.

%%% figure 2 %%%
\begin{figure}[!]
        \includegraphics[width=1.0\columnwidth]{../figures/figure_2.pdf}
        \caption{\label{fig:fig2} \textbf{Entropy measurement for a single spin $\frac{1}{2}$.} (a) Charge sensor data for $N=0 \rightarrow 1$ at two temperatures set by DC current through the QPC heater. (b) Transition width, $\Theta$, was linear in $T_{MC}$ above 100 mK, for $I_{heat}=0$. Lever arm $\alpha$ is calculated by fitting a straight line to this region.  (c) Lock-in measurement of $\delta G_{sens}$ with $\delta T = \SI{32}{\milli\kelvin}$, determined from the calibration in panel (d). Fits to $\delta G_{sens}$ (Eq.~\ref{eqn:dg-sens}) are shown with $\Delta S / k_B$  as a free parameter (solid) and fixed at $\Delta S / k_B=0$ (dashed). (d) $\Theta$ grows with DC current through the QPC heater. A fit to $T = a\sqrt{T^{2}_{MC}+b I^2_{heat}R_{QPC}}$ is used to convert between $I_{heat}$ and $\delta T$\cite{Mittal1996}. (e)  Entropy measurements were independent of the magnitude of $I_{heat}$ oscillations over a large range. The top axis indicates the corresponding magnitude of $\delta T$, while the right axis shows the entropy signal converted to a gate voltage shift per unit temperature.}
\end{figure}

%To proceed with the entropy measurement,
The dot was tuned such that the source was weakly tunnel coupled to the reservoir with the drain closed.  The conductance of the charge sensor was tuned to $G_{sens}{\sim}e^2/h$, where it was most sensitive to charge on the dot.  The addition of the first electron to the dot was marked by a decrease in $G_{sens}$ that is well described as a thermally-broadened two-level transition (Fig.~\ref{fig:fig2}a):
%
\begin{align}
\label{eqn:g-sens}
        G_{sens}(V_p,\Theta) &= G_0 \tanh\left(\frac{V_p - V_{mid}(\Theta)}{2\Theta}\right)  \\
                        &\quad + G_1 V_{G} + G_2 \nonumber
\end{align}
%
where $G_0$ quantifies the sensor sensitivity, $\Theta = \frac{k_B T}{\alpha e}$ is the thermal broadening (expressed in units of gate voltage), $\alpha\equiv d \mu_{N}/d V_p$ is the lever arm,  $G_1$ reflects the cross capacitance between the charge sensor and plunger gate, and $G_2$ is an offset. Figure~\ref{fig:fig2}a shows two such transition curves with thermal broadening set by the current, $I_{heat}$ through the heater. For $I_{heat}=0$, $\Theta$ followed $T_{MC}$ down to \SI{<100}{\milli\kelvin} (Fig.~\ref{fig:fig2}b), validating the approximation of thermal broadening used throughout this experiment.

The data in Fig.~\ref{fig:fig2}c, and corresponding fits, illustrate a measurement of $\Delta S_{0\rightarrow 1}$ across the $0 \rightarrow 1$ electron transition. The lock-in measurement of $\delta G_{sens}$ due to temperature oscillations $\delta T$ yields the characteristic peak-dip structure seen in Fig.~\ref{fig:fig2}c, corresponding the difference between the two $G_{sens}$ curves in Fig.~\ref{fig:fig2}a. The expected lineshape of such a curve is $\delta G_{sens} = \frac{\partial G_{sens}}{\partial T} \delta T$, with $G_{sens}$ defined by Eq. \ref{eqn:g-sens}.  This lineshape depends in a simple way on $\Delta S$, recognizing (via Eq.~\ref{eqn:max}) that $\frac{\partial V_{mid}}{\partial \Theta}=\frac{1}{k_B}\frac{\partial \mu}{\partial T} =\frac{1}{k_B}\Delta S_{N-1\rightarrow N}$:
%
\begin{align}
\label{eqn:dg-sens}
        \delta G_{sens}(V_p, \Theta) &\propto -\delta T \left[ \frac{V_p - V_{mid}(\Theta)}{2\Theta} - \frac{\Delta S}{2k_B} \right]\times \\
        				      &\quad\cosh^{-2}\left(\frac{V_p - V_{mid}(\Theta)}{2\Theta}\right) + const. \nonumber
\end{align}
%

\noindent As expected  from Figs.~\ref{fig:fig1}b,c, $\delta G_{sens}(V_p)$ is antisymmetric around $V_{mid}$ for $\Delta S=0$, and asymmetric for $\Delta S\neq 0$.  A fit of the data in Fig.~\ref{fig:fig2}c to Eq.~\ref{eqn:dg-sens} yields $\Delta S_{0\rightarrow 1}=(1.02 \pm 0.03) k_B \ln{2}$, closely matching the expected $\Delta S_{0 \rightarrow 1} = S_1 - S_0 =k_B\ln{2}$ for transitions between an empty dot ($S_0=0$) and the two-fold degenerate 1-electron state ($d_1=2$) with entropy $S_1=k_B \ln{2}$.

It is important to note that extracting $\Delta S$ from fits to Eq.~\ref{eqn:dg-sens} can be done just from the asymmetry of the lineshape, with no calibration of measurement parameters (such as $\delta T$ or the lever arm $\alpha$) required.  We can, however, estimate $\delta T$ and $\alpha$ by measuring $T(I_{heat})$, using a DC $I_{heat}$ to raise the reservoir temperature and fitting to Eq.~\ref{eqn:g-sens}  (Fig.~\ref{fig:fig2}d); the calibration holds for AC currents below 100 Hz.  Measurements of $\Delta S$ remained constant over a broad range of $\delta T$ (Fig.~\ref{fig:fig2}e), as expected for temperatures low enough not to excite orbital degrees of freedom on the dot.

%%% figure 3 %%%
\begin{figure}
        \includegraphics[width=1.0\columnwidth]{../figures/figure_3.pdf}
        \caption{\label{fig:fig3} \textbf{Magnetic field dependence.} (a) Changes in entropy for $N=0 \rightarrow 1$ and $2 \rightarrow 3$ transitions, overlaid to highlight similar behaviors.  Each data point corresponds to a single $\delta G_{sens}(V_p)$ fit; multiple scans are carried out at various parallel magnetic fields.   (b) and (c) Characteristic $\delta G_{sens}$ traces from which the data in (a) were extracted. The two data points corresponding to (b) and (c) are shown as large markers in (a). (e) Bias spectroscopy data for the $N=0 \rightarrow 1$ transition. Dashed line at $V_{SD}$ = \SI{700}{\micro\electronvolt} shows where data in (f) are taken. (f) Fixed bias data showing fits to Zeeman splitting of the ground state (dashed lines) from which $g = -0.42 \pm 0.01$ is extracted.}
\end{figure}
